---
title: "Poverty vs. Education Report"
author: "Liam Lindhe-Johan, Maryory Ajpop, Anand Puri, Beem Chivatxaranukul"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction

In researching a topic for this project the group decided they wanted to focus not on something that could be arbitrarily related, but a subject we knew must in some way be correlated. We wanted to tackle an issue that many are aware of, but often overlook due to how abstract of a concept it can be. The group decided on level of education versus amount of poverty in a location in the united states because we wanted a better understanding of where and how divisions in the educational field begin to arise. This subject is crucially important for the future of both education and economics in the United States as the relationship between these variables can have impacts that span generations.

We already know that at a baseline there is a correlation between income and education. This is due to several widely known and understood factors such as segregation, redlining, discrimination, the list goes on. The group already knew about these factors before deciding on our final data sets, and that was part of the reason we all felt it important to do our project on this subject in particular. While we all understood that this problem existed none of us knew where it was worst or to what extent it is still prevalent, and we felt that if we had these questions it would be likely that others did as well.

The questions that our group posed to focus our research and our project design process were:

1.  how does the education of an area correlate to the poverty of that area?

2.  how do different levels of education percentage correlate to poverty of a county?

3.  does a statewide higher level of education correlate to a statewide lower poverty rate?

4.  does completing a high school education significantly lower the chance of poverty?

In answering these questions we wanted to further understand the extent to which education affects the poverty of a location and if a more educated population truly leads to a lower rate of poverty.

This topic is incredibly important for the future of the United States and is intrinsically tied to almost every person's life to some extent. Recognizing how prominent and impactful these factors are in the overall poverty of the USA is a crucial step in the improvement of these issues, no matter what educational or economic background you originate from. These issues have affected all of us either directly or indirectly throughout the course of our lives and for us as a country to begin healing some of this damage we must first understand how and where it exists. This report will attempt to shine a light on the places in which the problems of education and poverty plague our country the most, and we as a group will seek to educate ourselves and others on the importance of these issues.

# Data

The data sets we used were:

# 

```{r}
library(tidyverse)
eduStats <- read.csv("Data/Education.csv")
tibble(eduStats)

povEst <- read.csv("Data/PovertyEstimates.csv")
tibble(povEst)
```

# 

# 

While they don't show up well unformatted in R, they functioned well for our purposes as they aligned with each other quite nicely and gave us more than the amount of information we were initially looking for.

Both of our data sets came from the Economic Research Service which is a part of the federal government with the goal of documenting important statistics such as these.

As this is a public and freely accessible arm of the federal government, anyone with an internet connection and the means to store the files can access both of these data sets. The data does not have any specific restrictions on it and is free to be used by anyone for essentially any purpose.

The data sets are fairly broad with entries on multiple facets of both education and poverty in all of the available counties in each state of the United States. The population of the data sets were clearly the residents of the USA and their education amount and poverty level. The sample for both of these data sets is the responses they received from their online surveys and the contents of the immense library of past survey responses from the rest of the government.

There were a few minor issues with the data such as formatting and misleading column labels, these are fairly visible from the earlier visualizations. Before we could begin using the data we had to overhaul the categorization and formatting system both data sets originally had. The worse of the two issues was absolutely the column titles, especially in the poverty data set in which they are entirely unviewable without a secondary cheat sheet defining each column individually.

#Results
We opted to utilize the formatted data to generate a set of four distinct graphical representations detailing and projecting the relationship between education levels and poverty rates. These visualizations included line graphs delineating the connections between different educational levels and poverty rates within populations, maps illustrating the distribution of poverty percentages across various counties in correlation with educational attainment, as well as a bar graph portraying the distribution of education levels among individuals experiencing poverty.
```{r}
edupov %>% 
  ggplot(aes(totalEducation, PCTPOVALL_2021)) +
  geom_point(col = "black") +
  geom_smooth(method = "lm", col = "pink2") +
  theme_minimal() +
  labs(
    x = "% of People With Highschool Education or Better",
    y = "% of People in Poverty",
    title = "Effect of Education on Poverty"
  )


edupovlong %>% 
  group_by(educationLevel) %>% 
  ggplot(aes(percent, PCTPOVALL_2021, col = educationLevel)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(
    x = "Education Level",
    y = "Percent in Poverty"
  )
edupov$highschoolOrLower <- 100 - edupov$Percent.of.adults.with.a.bachelor.s.degree.or.higher..2017.21 -
  edupov$Percent.of.adults.completing.some.college.or.associate.s.degree..2017.21

edupov %>% 
  ggplot(aes(highschoolOrLower, PCTPOVALL_2021)) +
  geom_point() +
  geom_smooth(method = "lm", col = "pink2") +
  labs(
    x = "Highschool Degree or Less",
    y = "Rate of Poverty in County",
    title = "Likelyhood of Poverty Based on Degree"
  ) +
  theme_minimal()
```


In our examination of the line graphs, a correlation emerged for individuals with a high school degree or lower which actually indicated a direct relationship between lower educational attainment and higher poverty rates. On the other hand, those with a college degree or higher exhibited an inverse relationship, showing decreasing poverty rates as the population of those with college degrees increased.

The geographic insights collected from the maps revealed another interesting pattern, where counties with lower percentages of educated individuals concurrently exhibited higher percentages of poverty. Through the 2 different maps of education level and poverty levels, counties where the education level was generally high showed lower frequencies of poverty. This spatial correlation blueprinted the intricate relationship between educational opportunities and economic well-being.
```{r}
chanceOfPoverty %>% 
  ggplot(aes(eduStages, eduAverages, fill = eduStages)) +
  geom_col() +
  theme_minimal() +
  labs(
    x = "Stages of Education",
    y = "Percentage of US Population in Poverty",
    title = "Education Levels of People in Poverty",
  ) +
  guides(fill = FALSE)
```

Final insights were derived from the bar graph, which communicated the distribution of levels of education among those in poverty. Through research conducted, there was an estimated 12% of people in the United States that were considered to be in poverty. Through this bar graph, it can be seen that around 7% of that 12% belonged to the group of people who had only a high school degree or less. The remaining percentages were shared between those with an associates degree and those with a full bachelor's degree or higher, with the former taking up a bit more of the share than the latter.






# Summary

During our work on this project, the group gained an important understanding of the true impact of the data we were studying. As we analyzed and processed our data sets the power of what we were creating became increasingly more evident. The relationships between these variables display a multitude of shortcomings within our country and an understanding of them is absolutely necessary if we do hope to solve these issues. Knowing what we know now about the direct impact education has on the poverty of an area we can look to further research on how and why these relationships exist in the first place. While the topic we chose is a fairly well-known relationship, many citizens of the United States are unaware of just how proportional these two variables are to each other. With the creation of our visualizations, we wanted to demonstrate a clear understanding of the data that could reach and be understood by as many as possible. With our project, we hoped to decrease barriers to data and create a framework for a future without this glaring problem in the United States.
